Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-09-22 manuscript/climate_sensitivity_draft.Rmd
finished reviewing/ condensing pass

Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**).
The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019].
Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**). 
There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020]. 
Drought tolerance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$).
This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
The directions of these responses were consistent across droughts (Table S8), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S8-S9), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. 
These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years [Fig. **1a**, @helcoski_growing_2019] and had substantial but short-lived impacts on tree growth (Fig. **1**). 
These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], but were modest compared to the so-called “megadroughts” that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019; @clark_impacts_2016]. 
Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). 
The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. 
Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits; Fig. **4**) had increased growth (Fig. **1b**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. 
Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. 1), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).

@@

Concerted measurement of tree-rings and leaf drought tolerance traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of drought tolerance traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig. **4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; @zhu_leaf_2018], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type. Further, it adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests. The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that do not form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@christoffersen_linking_2016; @powell_differences_2017].

>>>>

Tree height, microenvironment, and leaf drought tolerance traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**).
Taller trees had greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **2**), which is one plausible mechanism for their lower drought resistance, recovery, and resilience (Fig. **4**). 
There was no evidence that greater availability of, or access to, soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments.
The negative effect of height on $Rt$ held after accounting for species' traits, which is consistent with recent work finding height had a stronger influence on mortality risk than forest type during drought [@stovall_reply_2020]. 
Drought tolerance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf drought tolerance traits ($PLA_{dry}$, $\pi_{tlp}$).
This is the first study to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought.
The directions of these responses were consistent across droughts (Table S8), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S8-S9), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. 
These findings advance our knowledge of the factors that make trees vulnerable to stem growth declines during drought and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years [Fig. **1a**, @helcoski_growing_2019] and had substantial but short-lived impacts on tree growth (Fig. **1**). 
These droughts were classified as severe ($PDSI$ < -3.0; 1977) or extreme ($PDSI$ < -4.0; 1966, 1999) at our site and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], but were modest compared to the so-called “megadroughts” that have triggered massive tree die-off in other regions [e.g., @allen_global_2010; @stovall_tree_2019; @clark_impacts_2016]. 
Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S1**), severely stressed a larger portion of trees (Fig. **1b**). 
The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. 
Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion (e.g., short understory trees, species with drought resistant traits) had increased growth (Figs. **1b, 4**), consistent with prior observations that smaller trees can exhibit increased growth rates during drought [@bennett_larger_2015]. 
Growth rebounded strongly following the droughts, on average exceeding pre-drought growth rates (Fig. 1), particularly for shorter trees and species with drought-tolerant traits (Figs. **3-4**).
It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth (e.g., stand dynamics), that our best models characterize only a modest amount of variation in $Rt$, $Rc$, and $Rs$: 11-18% for all droughts combined, and 13-30% for individual droughts (Tables S8-S11).

@@

Concerted measurement of tree-rings and leaf drought tolerance traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of drought tolerance traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Fig. **4**; Table 1) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@marechaux_drought_2015; @fletcher_evolution_2018; @medeiros_extensive_2019; @simeone_coupled_2019;  @rosas_adjustments_2019; @zhu_leaf_2018], and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional type. Further, it adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models [@powell_differences_2017]. Because both $PLA_{dry}$ and $\pi_{tlp}$ can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across diverse forests. The importance of predicting drought responses from species traits increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for diverse tropical forests where most species do not form annual rings [but see @schongart_dendroecological_2017 for a review of progress in tropical dendroecology]. A full linkage of drought tolerance traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts [@christoffersen_linking_2016; @powell_differences_2017].
